Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2025-07-06 manuscript/reviews/ResponseReview1.Rmd

> 8. Table 2: It could be helpful to show an example of a trial where an event never
occurs, to help the reader connect what is written in the text (line 325-327) to
what appears in the table.

Response: We added data visualizations in section A of the Supplemental Material.

> 9. Figure 4: Why is there no bin 40?

Response: Because no responses occurred in the first bin with endpoint 40 ms. However, we added this bin back in Figure 2 (previously 4) of the revised ms.

@@

> 10. Line 423: The phrase “early responses” is confusing as it implies temporal
ordering of the responses (i.e., early vs. late trials). I’d suggest using “fast
response” or “rapid responses”, or some other phrase along those lines.

Response: We replaced early with fast on page 22 of the revised ms.

@@

> 11. Is the discussion of reference and index coding (lines 457 – 464) needed? In the
introduction, the authors state that knowledge of regression is assumed so I am
not sure this is necessary to include, but I defer to the authors’ judgment.

Response: Index coding is not used a lot in experimental psychology so we included a short description.

@@

> 12. Figure 6: I only see one width of credible interval for each line on this plot—
please clarify if you are plotting 80% or 95% credible intervals.

Response: We corrected this (Figure 4 in the revised ms.).

> 13. Line 536: To me, “subject-specific” and “marginal” mean the opposite things.
That is, subject-specific is an effect that is conditional on an individual while a
“marginal” effect has been marginalized/averaged over individuals. Could you
please clarify your use of these terms here?

Response: We feel there is no real consensus in the literature about the use of the term marginal. We thus removed this term in the revised ms. But we kept the term in tutorials 2a and 2b where we base our use of the terms on this cited reference: Heiss (2021) - https://www.andrewheiss.com/blog/2021/11/10/ame-bayes-re-guide/#average-marginal-effects.

@@

> There are many abbreviations that were difficult to follow. EHA is arguably standard, but SAT (for speed/accuracy trade-off) is less familiar, and may therefore benefit from not being an abbreviation. Similarly referring to hazard function as h(t), conditional accuracy function as c(t) was less clear than just saying hazard function and conditional accuracy function.

Response: Although SAT is rather standard in cognitive experimental psychology, we tried to limit the use of abbreviations in main body of the revised ms.

> It would be helpful if the research questions that can be addressed by the methods presented are clearly presented in each section. I think that currently they are getting lost in the tutorial.

Response: We now discuss general research questions on page 10 of the revised ms.

We did not add more information about possible research questions in each section of Tutorial 2a, because we feel that (a) research questions are highly dependent on the field of study (perception, behavioral control, memory), (b) this would make the tutorial even longer, and (c) the knowledge of experimental psychologists on regression modeling should be large enough for them to link the model structure to possible research questions, once they have read the paper and worked through the code examples.

@@

> It wasn’t clear how section 3.1 added value to the tutorial

Response: Section 3 has been removed.

> Section 3.2 – it would be helpful if there was clarification in the text on when it would not be possible to include varying intercepts and slopes (i.e., at least clarifying that random intercepts are a requirement, but sometimes random slopes are removed based on model convergence issues – however, this also begs the question of whether a model should even be presented if it does not converge when it includes random slopes…)

Response: We do not discuss frequentist approaches in the main text anymore (to save space) and only mention that the reader can consult Tutorials 3a and 3b. We agree that a model should not be presented if it does not converge when it includes random slopes. However, to save space, and because this is an active area of research, we do not attempt to clarify when it would be possible to include varying intercepts and/or slopes.

### 2025-07-06 manuscript/reviews/ResponseReview1.tex

\begin{quote}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{11}
\tightlist
\item
  Figure 6: I only see one width of credible interval for each line on
  this plot--- please clarify if you are plotting 80\% or 95\% credible
  intervals.
\end{enumerate}
\end{quote}

Response: We corrected this (Figure 4 in the revised ms.).

\begin{quote}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{12}
\tightlist
\item
  Line 536: To me, ``subject-specific'' and ``marginal'' mean the
  opposite things. That is, subject-specific is an effect that is
  conditional on an individual while a ``marginal'' effect has been
  marginalized/averaged over individuals. Could you please clarify your
  use of these terms here?
\end{enumerate}
\end{quote}

Response: We feel there is no real consensus in the literature about the
use of the term marginal. We thus removed this term in the revised ms.
But we kept the term in tutorials 2a and 2b where we base our use of the
terms on this cited reference: Heiss (2021) -
\url{https://www.andrewheiss.com/blog/2021/11/10/ame-bayes-re-guide/\#average-marginal-effects}.

### 2025-07-06 manuscript/manuscript.Rmd
update ms.

---
title             : "Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows"
shorttitle        : "A tutorial on event history analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
#    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
#      - "Conceptualization"
#      - "Writing - Original Draft Preparation"
#      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
#    role:
#      - "Conceptualization"
#      - "Writing - Review & Editing"
#      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  Social Brain Sciences lab, Department of Humanities, Social and Political Sciences (D-GESS).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times and saccade latencies form a cornerstone of experimental psychology, and have had a widespread impact on our understanding of human cognition. However, the orthodox method for analyzing such data -- comparing means between conditions -- is known to conceal valuable information about the timeline of psychological effects, such as their onset time and how they evolve with increasing waiting time. The ability to reveal finer-grained, "temporal states" of cognitive processes can have important consequences for theory development by qualitatively changing the key inferences that are drawn from psychological data. Luckily, well-established analytical approaches, such as event history analysis (EHA), are able to evaluate the detailed shape of time-to-event distributions, and thus characterize the time course of psychological states. One barrier to wider use of EHA, however, is that the analytical workflow is typically more time-consuming and complex than orthodox approaches. To help achieve broader uptake of EHA, in this paper we outline a set of tutorials that detail one distributional method known as discrete-time EHA. We touch upon several key aspects of the workflow, such as how to process raw data and specify regression models, and we also consider the implications for experimental design. We finish the article by considering the benefits of the approach for understanding psychological states, as well as the limitations and future directions of this work. Finally, the project is written in R and freely available, which means the approach can easily be adapted to other data sets.  
keywords          : "response times, event history analysis, Bayesian multilevel regression models, experimental psychology, cognitive psychology"
wordcount         : "10131 (body) + 1709 (references) + 3473 ( body supplemental material) + 393 (refs suppl. mat.)"
bibliography      : ["extrareferences2.bib", "r-references.bib"]
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man, donotrepeattitle"
#appendix          : SupplementaryMaterial.Rmd
output            : papaja::apa6_pdf
# output            : papaja::apa6_word
editor_options: 
  chunk_output_type: console
header-includes:
  - \raggedbottom
---

@@

(ref:descr-fig1-caption) Simulated data showing mean performance versus distributional (EHA/SAT) analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms. The first bin is (0,100], the last bin is (900,1000]. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2.1.2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B).

@@

EHA is a class of statistical approaches to study the occurrence and timing of events, such as disease onset, marriages, arrests, and job terminations [@allisonSurvivalAnalysisUsing2010]. In this section, we want to provide an intuition regarding how EHA works in general, as well as in the context of experimental psychology. For those who want more detailed treatment of EHA and/or regression equations, we refer the reader to several excellent textbooks on these topics [@singerAppliedLongitudinalData2003; @allisonSurvivalAnalysisUsing2010; @gelmanRegressionOtherStories2020; @winterStatisticsLinguistsIntroduction2019]. We supply a few regression equations in section E of the Supplemental Material.

@@

Third, the width of each time bin will need to be determined. 
For instance, in Figure 1B we chose 100 ms in an arbitrary manner.
In reality, however, bin width will need to be set by considering a number of factors simultaneously.
The optimal bin width will depend on (a) the length of the observation period in each trial, (b) the rarity of event occurrence, (c) the number of repeated measures (or trials) per condition per participant, and (d) the shape of the hazard function.
Finding an appropriate bin width in a given user case before fitting models will require testing a number of options, when calculating and plotting the descriptive statistics (see section 3.1). The goal is to find the smallest bin width that is supported by the amount of data available. Based on our experience, a bin width of 40 ms is a good starting value when the number of repeated measures is 100 or less. Too small bin widths will result in erratic hazard functions as many bins will have no events, and thus hazard estimates of zero. Interestingly, the time bins do not need to have the same width. For example, @panisHowCanWe2020 used larger bins towards the end of the observation period, as fewer events occurred there.

@@

Our data wrangling procedures serve two related purposes. First, we want to summarise and visualise descriptive statistics using a life table. A life table includes for each time bin, the risk set (i.e., the number of trials that are event-free at the start of the bin), the number of observed events, and the estimates of the discrete-time hazard probability h(t), survival probability S(t), probability mass P(t), possibly the conditional accuracy ca(t), and their estimated standard errors (se). The definitions of these quantities are provided in section B of the Supplemental Material.

@@

Figure 4 displays the discrete-time hazard, survivor, conditional accuracy, and probability mass functions for each prime condition for participant 6. By using discrete-time hazard functions of event occurrence -- in combination with conditional accuracy functions for two-choice tasks -- one can provide an unbiased, time-varying, and probabilistic description of the latency and accuracy of responses based on all trials of any RT data set.

(ref:descr-plot4-caption) Estimated discrete-time hazard (h), survivor (S), conditional accuracy (ca) and probability mass (P) functions for participant 6. Vertical dotted lines indicate the estimated median RTs. Error bars represent +/- 1 standard error of the respective proportion.

```{r eha-plot, fig.cap = "(ref:descr-plot4-caption)", out.width="90%"}
knitr::include_graphics("../Tutorial_1_descriptive_stats/figures/Plot_for_subject6_PanisSchmidt.png")
```

@@

At this point, we have calculated and plotted the descriptive statistics for each type of prime stimulus in EHA/SAT. As we will show in later Tutorials, statistical models for hazard and conditional accuracy functions can be implemented as generalized linear mixed regression models predicting event occurrence (1/0) and conditional accuracy (1/0) in each bin of a selected time window for analysis. But first we consider calculating the descriptive statistics for within-subject designs with two independent variables.

@@

The middle column of Supplementary Figure 3 (section F of the Supplemental Material) shows six examples of prior distributions for an intercept on the logit and/or cloglog scales. While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B of Supplementary Figure 3 show that specifying such distributions on the logit and cloglog scales actually leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1.

@@

To make causal inferences from the parameter estimates in model M1i, we first plot the densities of the draws from the posterior distributions of its population-level parameters in Figure 5, together with point (median) and interval estimates (80% and 95% credible intervals). A credible interval is a range of values that contains a parameter's true value with a specified probability, given the observed data and model.

(ref:descr-plot5-caption) Medians and 80/95% credible intervals of the posterior distributions of the population-level parameters of model M1i.

```{r plot-fixed-effects, fig.cap = "(ref:descr-plot5-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_postdistr.png")
```

Because the parameter estimates are on the cloglog-hazard scale, we can ease our interpretation by plotting the expected value of the posterior predictive distribution -- the predicted hazard values -- at the population level (Figure 6A), and for each participant in the data set (Figure 6B).

(ref:descr-plot6-caption) Point (median) and 80/95% credible interval summaries of the hazard estimates (expected values of the draws from the posterior predictive distributions) in each time bin at the population level (A), and for each participant (B).

```{r plot-pred-grand-effects, fig.cap = "(ref:descr-plot6-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_pred_combined.png")
```

As we are actually interested in the effects of congruent and incongruent primes, relative to the blank prime condition, we can construct two contrasts (congruent-blank, incongruent-blank), and plot the posterior distributions of these contrast effects, both at the population level (Figure 7A) and at the participant level (Figure 7B).

(ref:descr-plot7-caption) Point (mean) and 80/95% credible interval summaries of estimated differences in hazard in each time bin at the population level (A), and for each participant (B).

```{r plot-grand-ame-effects, fig.cap = "(ref:descr-plot7-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ame_combined.png")
```

@@

If we want to estimate the population-level effect of prime type on hazard, we can base our conclusion on the credible Intervals (CrIs) in Figure 4A. The contrast "congruent minus blank" was estimated to be 0.09 hazard units in bin (200,240] (95% CrI = [0.02, 0.17]), and 0.14 hazard units in bin (240,280]) (95% CrI = [0.04, 0.25]). For the other bins, the 95% credible interval contained zero.
The contrast "incongruent minus blank" was estimated to be 0.09 hazard units in bin (200,240] (95% CrI = [0.01, 0.21]), -0.19 hazard units in bin (320,360] (95% CrI = [-0.31, -0.06]), -0.27 hazard units in bin (360,400] (95% CrI = [-0.45, -0.04]), and -0.23 hazard units in bin (400,440] (95% CrI = [-0.40, -0.03]). For the other bins, the 95% credible interval contained zero.

@@

If we want to focus more on inter-individual differences, we can study the subject-specific hazard functions in Figure 4B. Note that three participants (1, 2, and 3) show a negative difference for the contrast "congruent minus incongruent" in bin (360,400] -- subject 2 also in bin (320,360].

@@

Interestingly, all subjects show a tendency in their mean difference (congruent minus blank) to "dip" around that time (Figure 4B). Therefore, future modeling efforts could incorporate the trial number into the model formula, in order to also study how the effects of prime type on hazard change on the long experiment-wide time scale, next to the short trial-wide time scale. In Tutorial_2a.Rmd we provide a number of model formulae that should get you going.

@@

To make inferences from the parameter estimates in model M1i_ca, we first plot the densities of the draws from the posterior distributions of its population-level parameters in Figure 8, together with point (median) and interval estimates (80% and 95% credible intervals).

(ref:descr-plot8-caption) Medians and 80/95% credible intervals of the posterior distributions of the population-level parameters of model M1i_ca. ca = conditional accuracy.

```{r plot-ca-fixed-effects, fig.cap = "(ref:descr-plot8-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ca_postdistr.png")
```

Because the parameter estimates are on the logit-ca scale, we can ease our interpretation by plotting the expected value of the posterior predictive distribution -- the predicted conditional accurcies -- at the population level (Figure 9A), and for each participant in the data set (Figure 9B).

(ref:plot9-caption) Point (median) and 80/95% credible interval summaries of the conditional accuracy estimates (expected values of the draws from the posterior predictive distributions) in each time bin at the population level (A), and for each participant (B).

```{r plot-pred-grand-effects-ca, fig.cap = "(ref:plot9-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ca_pred_combined.png")
```

As we are actually interested in the effects of congruent and incongruent primes, relative to the blank prime condition, we can construct two contrasts (congruent-blank, incongruent-blank), and plot the posterior distributions of these contrast effects at the population level (Figure 10A) and for each participant (Figure 10B).

(ref:descr-plot10-caption) Point (mean) and 80/95% credible interval summaries of estimated differences in conditional accuracy in each time bin at the population level (A), and for each participant (B).

```{r plot-ca-grand-ame-effects, fig.cap = "(ref:descr-plot10-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ca_ame_combined.png")
```

Based on Figure 10A we see that on the population level congruent primes have a positive effect on the conditional accuracy of emitted responses in time bins (200,240], (240,280], (280,320], and (320,360], relative to the estimates in the baseline condition (blank prime; red dashed lines in Figure 10A). 
Incongruent primes have a negative effect on the conditional accuracy of emitted responses in the first time bins, relative to the estimates in the baseline condition.

@@

We then took parameters from this model and used them to create a single dataset with 200 trials per condition for 10 individual participants. The raw data and the simulated data are plotted in Figure 12 and show quite close correspondence, which is re-assuring. But, this is only one dataset. What we really want to do is simulate many datasets and vary parameters of interest, which is what we turn to in the next section.

@@

Summary results from Simulation 1 are shown in Figure 13A.
Figure 13A depicts statistical "power" as calculated by the percentage of lower-bound 95% confidence intervals that exclude zero when the difference between prime condition is calculated (congruent - incongruent). 
In other words, what fraction of the simulated datasets generated an effect of prime that excludes the criterion mark of zero.
We are aware that "power" is not part of a Bayesian analytical workflow, but we choose to include it here, as it is familiar to most researchers in experimental psychology.

@@

In Simulation 2, we varied the effect size between a different range of values (0.5, 0.4, 0.3), which correspond to a 50%, 60% and 70% reduction in hazard value as a function of prime condition.
In addition, we varied the number of participants per experiment between 10, 15, and 20 participants.
Given that trial count per condition made little difference to power in Simulation 1, we fixed trial count at 200 trials per condition in Simulation 2.
Summary results from Simulation 2 are shown in Figure 13B.
A summary of these power calculations might be as follows (trial count = 200 per condition in all cases):

>>>>

---
title             : "Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows"
shorttitle        : "A tutorial on event history analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
#    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
#      - "Conceptualization"
#      - "Writing - Original Draft Preparation"
#      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
#    role:
#      - "Conceptualization"
#      - "Writing - Review & Editing"
#      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  Social Brain Sciences lab, Department of Humanities, Social and Political Sciences (D-GESS).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times and saccade latencies form a cornerstone of experimental psychology, and have had a widespread impact on our understanding of human cognition. However, the orthodox method for analyzing such data -- comparing means between conditions -- is known to conceal valuable information about the timeline of psychological effects, such as their onset time and how they evolve with increasing waiting time. The ability to reveal finer-grained, "temporal states" of cognitive processes can have important consequences for theory development by qualitatively changing the key inferences that are drawn from psychological data. Luckily, well-established analytical approaches, such as event history analysis (EHA), are able to evaluate the detailed shape of time-to-event distributions, and thus characterize the time course of psychological states. One barrier to wider use of EHA, however, is that the analytical workflow is typically more time-consuming and complex than orthodox approaches. To help achieve broader uptake of EHA, in this paper we outline a set of tutorials that detail one distributional method known as discrete-time EHA. We touch upon several key aspects of the workflow, such as how to process raw data and specify regression models, and we also consider the implications for experimental design. We finish the article by considering the benefits of the approach for understanding psychological states, as well as the limitations of this work. Finally, the project is written in R and freely available, which means the approach can easily be adapted to other data sets.  
keywords          : "response times, event history analysis, Bayesian multilevel regression models, experimental psychology, cognitive psychology"
wordcount         : "10131 (body) + 1709 (references) + 3473 ( body supplemental material) + 393 (refs suppl. mat.)"
bibliography      : ["extrareferences2.bib", "r-references.bib"]
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man, donotrepeattitle"
#appendix          : SupplementaryMaterial.Rmd
output            : papaja::apa6_pdf
# output            : papaja::apa6_word
editor_options: 
  chunk_output_type: console
header-includes:
  - \raggedbottom
---

@@

(ref:descr-fig1-caption) Simulated single-subject data showing mean performance versus distributional (EHA/SAT) analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms. The first bin is (0,100], the last bin is (900,1000]. Two hundred trials were simulated in each condition. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2.1.2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B).

@@

EHA is a class of statistical approaches to study the occurrence and timing of events, such as disease onset, marriages, arrests, and job terminations [@allisonSurvivalAnalysisUsing2010]. In this section, we want to provide an intuition regarding how EHA works in general, as well as in the context of experimental psychology. For those who want more detailed treatment of EHA and/or regression equations, we refer the reader to several excellent textbooks on these topics [@singerAppliedLongitudinalData2003; @allisonSurvivalAnalysisUsing2010; @gelmanRegressionOtherStories2020; @winterStatisticsLinguistsIntroduction2019; @millsIntroducingSurvivalEvent2011]. We supply a few regression equations in section E of the Supplemental Material.

@@

Third, the width of each time bin will need to be determined. 
For instance, in Figure 1B we chose 100 ms in an arbitrary manner.
In reality, however, bin width will need to be set by considering a number of factors simultaneously.
The optimal bin width will depend on (a) the length of the observation period in each trial, (b) the rarity of event occurrence, (c) the number of repeated measures (or trials) per condition per participant, and (d) the shape of the hazard function.
Finding an appropriate bin width in a given user case before fitting models will require testing a number of options, when calculating and plotting the descriptive statistics (see section 3.1). The goal is to find the smallest bin width that is supported by the amount of data available. Based on our experience, a bin width of 50 ms is a good starting value when the number of repeated measures is 100 or less. Too small bin widths will result in erratic hazard functions as many bins will have no events, and thus hazard estimates of zero. Interestingly, the time bins do not need to have the same width. For example, @panisHowCanWe2020 used larger bins towards the end of the observation period, as fewer events occurred there.

@@

Our data wrangling procedures serve two related purposes. First, we want to calculate descriptive statistics for each condition in each individual using a life table. A life table includes for each time bin, the risk set (i.e., the number of trials that are event-free at the start of the bin), the number of observed events, and the estimates of the discrete-time hazard probability h(t), survival probability S(t), probability mass P(t), possibly the conditional accuracy ca(t), and their estimated standard errors (se). The definitions of these quantities are provided in section B of the Supplemental Material.

@@

Figure 2 displays the discrete-time hazard, survivor, conditional accuracy, and probability mass functions for each prime condition for participant 6. By using discrete-time hazard functions of event occurrence -- in combination with conditional accuracy functions for two-choice tasks -- one can provide an unbiased, time-varying, and probabilistic description of the latency and accuracy of responses based on all trials of any RT data set.

(ref:descr-plot2-caption) Estimated discrete-time hazard (h), survivor (S), conditional accuracy (ca) and probability mass (P) functions for participant 6. Vertical dotted lines indicate the estimated median RTs. Error bars represent +/- 1 standard error of the respective proportion.

```{r eha-plot, fig.cap = "(ref:descr-plot2-caption)", out.width="90%"}
knitr::include_graphics("../Tutorial_1_descriptive_stats/figures/Plot_for_subject6_PanisSchmidt.png")
```

@@

At this point, we have calculated and plotted the descriptive statistics for each type of prime stimulus. As we will show in later Tutorials, statistical models for hazard and conditional accuracy functions can be implemented as generalized linear mixed regression models predicting event occurrence (1/0) and conditional accuracy (1/0) in each bin of a selected time window for analysis. But first we consider calculating the descriptive statistics for within-subject designs with two independent variables.

@@

The middle column of Supplementary Figure 2 (section D of the Supplemental Material) shows six examples of prior distributions for an intercept on the logit and/or cloglog scales. While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B of Supplementary Figure 2 show that specifying such distributions on the logit and cloglog scales actually leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1.

@@

To make causal inferences from the parameter estimates in model M1i, we first plot the densities of the draws from the posterior distributions of its population-level parameters in Figure 3, together with point (median) and interval estimates (80% and 95% credible intervals). A credible interval is a range of values that contains a parameter's true value with a specified probability, given the observed data and model.

(ref:descr-plot3-caption) Medians and 80/95% credible intervals of the posterior distributions of the population-level parameters of model M1i.

```{r plot-fixed-effects, fig.cap = "(ref:descr-plot3-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_postdistr.png")
```

Because the parameter estimates are on the cloglog-hazard scale, we can ease our interpretation by plotting the expected value of the posterior predictive distribution -- the predicted hazard values -- at the population level (Figure 4A), and for each participant in the data set (Figure 4B).

(ref:descr-plot4-caption) Point (median) and 80/95% credible interval summaries of the hazard estimates (expected values of the draws from the posterior predictive distributions) in each time bin at the population level (A), and for each participant (B).

```{r plot-pred-grand-effects, fig.cap = "(ref:descr-plot4-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_pred_combined.png")
```

As we are actually interested in the effects of congruent and incongruent primes, relative to the blank prime condition, we can construct two contrasts (congruent-blank, incongruent-blank), and plot the posterior distributions of these contrast effects, both at the population level (Figure 5A) and at the participant level (Figure 5B).

(ref:descr-plot5-caption) Point (mean) and 80/95% credible interval summaries of estimated differences in hazard in each time bin at the population level (A), and for each participant (B).

```{r plot-grand-ame-effects, fig.cap = "(ref:descr-plot5-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ame_combined.png")
```

@@

If we want to estimate the population-level effect of prime type on hazard, we can base our conclusion on the credible Intervals (CrIs) in Figure 5A. The contrast "congruent minus blank" was estimated to be 0.09 hazard units in bin (200,240] (95% CrI = [0.02, 0.17]), and 0.14 hazard units in bin (240,280]) (95% CrI = [0.04, 0.25]). For the other bins, the 95% credible interval contained zero.
The contrast "incongruent minus blank" was estimated to be 0.09 hazard units in bin (200,240] (95% CrI = [0.01, 0.21]), -0.19 hazard units in bin (320,360] (95% CrI = [-0.31, -0.06]), -0.27 hazard units in bin (360,400] (95% CrI = [-0.45, -0.04]), and -0.23 hazard units in bin (400,440] (95% CrI = [-0.40, -0.03]). For the other bins, the 95% credible interval contained zero.

@@

If we want to focus more on inter-individual differences, we can study the subject-specific hazard functions in Figure 5B. Note that three participants (1, 2, and 3) show a negative difference for the contrast "congruent minus incongruent" in bin (360,400] -- subject 2 also in bin (320,360]. Interestingly, all subjects show a tendency in their mean difference (congruent minus blank) to "dip" around that time. Therefore, future modeling efforts could incorporate the trial number into the model formula, in order to also study how the effects of prime type on hazard change on the long experiment-wide time scale, next to the short trial-wide time scale. In Tutorial_2a.Rmd we provide a number of model formulae that should get you going.

@@

@@

To make inferences from the parameter estimates in model M1i_ca, we first plot the densities of the draws from the posterior distributions of its population-level parameters in Figure 6, together with point (median) and interval estimates (80% and 95% credible intervals).

(ref:descr-plot6-caption) Medians and 80/95% credible intervals of the posterior distributions of the population-level parameters of model M1i_ca. ca = conditional accuracy.

```{r plot-ca-fixed-effects, fig.cap = "(ref:descr-plot6-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ca_postdistr.png")
```

Because the parameter estimates are on the logit-ca scale, we can ease our interpretation by plotting the expected value of the posterior predictive distribution -- the predicted conditional accurcies -- at the population level (Figure 7A), and for each participant in the data set (Figure 7B).

(ref:plot7-caption) Point (median) and 80/95% credible interval summaries of the conditional accuracy estimates (expected values of the draws from the posterior predictive distributions) in each time bin at the population level (A), and for each participant (B).

```{r plot-pred-grand-effects-ca, fig.cap = "(ref:plot7-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ca_pred_combined.png")
```

As we are actually interested in the effects of congruent and incongruent primes, relative to the blank prime condition, we can construct two contrasts (congruent-blank, incongruent-blank), and plot the posterior distributions of these contrast effects at the population level (Figure 8A) and for each participant (Figure 8B).

(ref:descr-plot8-caption) Point (mean) and 80/95% credible interval summaries of estimated differences in conditional accuracy in each time bin at the population level (A), and for each participant (B).

```{r plot-ca-grand-ame-effects, fig.cap = "(ref:descr-plot8-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ca_ame_combined.png")
```

Based on Figure 8A we see that on the population level congruent primes have a positive effect on the conditional accuracy of emitted responses in time bins (200,240], (240,280], (280,320], and (320,360], relative to the estimates in the baseline condition (blank prime; red dashed lines in Figure 8A). 
Incongruent primes have a negative effect on the conditional accuracy of emitted responses in the first time bins, relative to the estimates in the baseline condition.

@@

We then took parameters from this model and used them to create a single dataset with 200 trials per condition for 10 individual participants. The raw data and the simulated data are plotted in Figure 9 and show quite close correspondence, which is re-assuring. But, this is only one dataset. What we really want to do is simulate many datasets and vary parameters of interest, which is what we turn to in the next section.

@@

Summary results from Simulation 1 are shown in Figure 10A.
Figure 10A depicts statistical "power" as calculated by the percentage of lower-bound 95% confidence intervals that exclude zero when the difference between prime condition is calculated (congruent - incongruent). 
In other words, what fraction of the simulated datasets generated an effect of prime that excludes the criterion mark of zero.
We are aware that "power" is not part of a Bayesian analytical workflow, but we choose to include it here, as it is familiar to most researchers in experimental psychology.

@@

In Simulation 2, we varied the effect size between a different range of values (0.5, 0.4, 0.3), which correspond to a 50%, 60% and 70% reduction in hazard value as a function of prime condition.
In addition, we varied the number of participants per experiment between 10, 15, and 20 participants.
Given that trial count per condition made little difference to power in Simulation 1, we fixed trial count at 200 trials per condition in Simulation 2.
Summary results from Simulation 2 are shown in Figure 10B.
A summary of these power calculations might be as follows (trial count = 200 per condition in all cases):
